---
title: "Worksheet 2"
author: 'Goliath : Wenje Tu, Lea Bührer, ...'
date: "Spring Semester 2022"
output: pdf_document
# bibliography: biblio.bib
nocite: '@*'
subtitle: Foundations of Bayesian Methodology
papersize: a4
fontsize: 10pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Exercise 3 (Conjugate Bayes: analytical derivation)

Assumptions: 

$$
\begin{aligned}
y_1,\cdots,y_n\mid m  & \overset{i.i.d}{\sim} \mathcal{N}(m,\kappa^{-1}) \\
m & \: \sim \:\mathcal{N}(\mu,\lambda^{-1})
\end{aligned}
$$
The Likelihood is equal 
$$
\begin{aligned}
f(y_1,\cdots,y_n\mid m)
&= \prod_{i=1}^{n} f(y_i\mid m) \\
&= \prod_{i=1}^{n} \left(\sqrt{\frac{\kappa}{2\pi}}\exp\left(-\frac{\kappa}{2}(y_i-m)^2 \right) \right) \\
&= \left(\frac{\kappa}{2\pi} \right)^{\frac{n}{2}}\exp\left(-\frac{\kappa}{2}\sum_{i=1}^{n}(y_i-m)^2 \right)
\end{aligned}
$$
Prior density distribution of m: 
$$
f(m)=\sqrt{\frac{\lambda}{2\pi}}\exp\left(-\frac{\lambda}{2}(m-\mu)^2 \right)
$$
In a next step the likelihood and prior density function of m are multiplied: 
$$
\begin{aligned}
&\quad\quad f(y_1,\cdots,y_n\mid m)f(m) \\
&= \left(\frac{\kappa}{2\pi} \right)^{\frac{n}{2}}\exp\left(-\frac{\kappa}{2}\sum_{i=1}^{n}(y_i-m)^2 \right)
\cdot \sqrt{\frac{\lambda}{2\pi}}\exp\left(-\frac{\lambda}{2}(m-\mu)^2 \right) \\
&= \left(\frac{\kappa}{2\pi} \right)^{\frac{n}{2}}\left(\frac{\lambda}{2\pi}\right)^{\frac{1}{2}}
\exp\left(-\frac{\kappa}{2}\sum_{i=1}^{n}(y_i-m)^2-\frac{\lambda}{2}(m-\mu)^2 \right) \\
&= \left(\frac{\kappa}{2\pi} \right)^{\frac{n}{2}}\left(\frac{\lambda}{2\pi}\right)^{\frac{1}{2}}
\exp\left(-\frac{\kappa}{2}\sum_{i=1}^{n}(y_i^2-2y_im+m^2)-\frac{\lambda}{2}(m^2-2m\mu+\mu^2) \right) \\
&= \left(\frac{\kappa}{2\pi} \right)^{\frac{n}{2}}\left(\frac{\lambda}{2\pi}\right)^{\frac{1}{2}}
\exp\left(-\frac{\kappa}{2}\left(\sum_{i=1}^{n}y_i^2-2n\bar{y}m+n m^2\right)-\frac{\lambda}{2}(m^2-2m\mu+\mu^2) \right) \\
&= \left(\frac{\kappa}{2\pi} \right)^{\frac{n}{2}}\left(\frac{\lambda}{2\pi}\right)^{\frac{1}{2}}
\exp\left(-\frac{1}{2}\left((n\kappa+\lambda)m^2-2m(\kappa n\bar{y}+\lambda\mu)+\kappa\sum_{i=1}^{n}y_i^2+\lambda\mu^2 \right) \right) \\
&= \left(\frac{\kappa}{2\pi} \right)^{\frac{n}{2}}\left(\frac{\lambda}{2\pi}\right)^{\frac{1}{2}}
\exp\left(-\frac{n\kappa+\lambda}{2}\left(m^2-2m\frac{\kappa n\bar{y}+\lambda\mu}{n\kappa+\lambda}+\frac{\kappa\sum_{i=1}^{n}y_i^2+\lambda\mu^2}{n\kappa+\lambda} \right) \right) \\
&= \left(\frac{\kappa}{2\pi} \right)^{\frac{n}{2}}\left(\frac{\lambda}{2\pi}\right)^{\frac{1}{2}}
\exp\left(-\frac{n\kappa+\lambda}{2}\left(\left(m-\frac{\kappa n\bar{y}+\lambda\mu}{n\kappa+\lambda} \right)^2-\left(\frac{\kappa n\bar{y}+\lambda\mu}{n\kappa+\lambda}\right)^2+\frac{\kappa\sum_{i=1}^{n}y_i^2+\lambda\mu^2}{n\kappa+\lambda} \right) \right) \\
&= \underbrace{
\left(\frac{\kappa}{2\pi} \right)^{\frac{n}{2}}\left(\frac{\lambda}{2\pi}\right)^{\frac{1}{2}}
\exp\left(\frac{\kappa\sum_{i=1}^{n}y_i^2+\lambda\mu^2}{n\kappa+\lambda}-\left(\frac{\kappa n\bar{y}+\lambda\mu}{n\kappa+\lambda}\right)^2\right)
}_\text{constant}
\exp\left(-\frac{n\kappa+\lambda}{2}\left(m-\frac{\kappa n\bar{y}+\lambda\mu}{n\kappa+\lambda} \right)^2  \right) \\
\end{aligned}
$$
According to the Bayes formula, which has been rewritten in terms of densities, we get: 
$$
\begin{aligned}
f(m\mid y_1,\cdots, y_n)
&= \frac{f(y_1,\cdots,y_n\mid m)f(m)}{f(y_1,\cdots,y_n)} \\
&= \frac{f(y_1,\cdots,y_n\mid m)f(m)}{\int_{-\infty}^{\infty}f(y_1,\cdots,y_n\mid m)f(m)\textrm{d}m} \\
&= \frac{\exp\left(-\frac{n\kappa+\lambda}{2}\left(m-\frac{\kappa n\bar{y}+\lambda\mu}{n\kappa+\lambda} \right)^2  \right)}
{\int_{-\infty}^{\infty}\exp\left(-\frac{n\kappa+\lambda}{2}\left(m-\frac{\kappa n\bar{y}+\lambda\mu}{n\kappa+\lambda} \right)^2  \right)}  \\
&= \sqrt{\frac{n\kappa+\lambda}{2\pi}}\exp\left(-\frac{n\kappa+\lambda}{2}\left(m-\frac{\kappa n\bar{y}+\lambda\mu}{n\kappa+\lambda} \right)^2  \right) 
\end{aligned}
$$

$$
m\mid y_1,\cdots,y_n\sim \mathcal{N}
\left(\frac{\kappa n\bar{y}+\lambda\mu}{n\kappa+\lambda}, \frac{1}{n\kappa+\lambda}\right)
$$

Note:

$$
\int_{-\infty}^{\infty}e^{-a(x+b)^2}\textrm{d}x=\sqrt{\frac{\pi}{\alpha}}
$$

Source: [Gaussian Integral](https://en.wikipedia.org/wiki/Gaussian_integral)



### Exercise 4 (Conjugate Bayesian analysis in practice)


Assumptions: 

$$
\begin{aligned}
y_1,\cdots,y_n\mid m  & \overset{i.i.d}{\sim} \mathcal{N}(m,\kappa^{-1}) \\
m & \: \sim \:\mathcal{N}(\mu,\lambda^{-1}) 
\end{aligned}
$$

$$
\kappa=\frac{1}{900},\quad \mu=161, \quad \lambda=\frac{1}{70}
$$

#### 4(a)

```{r}
height <- c(166, 168, 168, 177, 160, 170, 172, 159, 175, 164, 175, 167, 164)
table(height)
```


```{r, out.width="50%", fig.align='center', fig.cap="Histogram of the height measurements\\label{fig:hist}"}
hist(height, breaks=length(height), freq=FALSE)
lines(density(height), col="red")
```
```{r, out.width="50%", fig.align='center', fig.cap="Q-Q Plot of the Measurements\\label{fig:qqhueght}"}

library(car)
qqPlot(height, main="Normal Q-Q Plot", ylab="Sample", xlab="Norm Quantiles")

# or alternatively without library(car)
qqnorm(height)
qqline(height)
```

In the Q-Q Plot it can be seen that all sample values lie within the area where normal distribution of the data can be assumed. Due to the fact that we only have 13 observations the 95%-CI has been calculated assuming a t-distribution. 


```{r}
## Sample size
sample.n <- length(height); sample.n

## Sample meam
sample.mean <- mean(height); sample.mean

## Sample standard deviation
sample.sd <- sd(height); sample.sd

## Sample standard error
sample.se <- sample.sd / sqrt(sample.n); sample.se
```

```{r}
## Significance level
alpha <- 0.05

## Degrees of freedom
df <- sample.n - 1

## t-score
t.score <- qt(p=alpha/2, df=df, lower.tail=FALSE); t.score
```

```{r}
## Contract the 95% confidence interval
lower.bound <- sample.mean - t.score * sample.se
upper.bound <- sample.mean + t.score * sample.se
print(c(lower.bound, upper.bound))
```

**Interpretation**: there is a probability of 95% that the true mean will fall into the interval $(164.6722, 171.4816)$.

#### 4(b)


```{r, out.width="50%", fig.align='center', fig.cap="Prior Distribution\\label{fig:prior}"}
## Plot of the Prior Distribution: 
m <- seq(101, 221)
plot(m, dnorm(m, mean=161, sd=sqrt(70)), type="l", ylab="density", col="red", 
     ylim=c(0,0.07), main="Prior vs Posterior", lwd=2)
legend("topright", legend=c("prior: N(161,70)"), 
       lty=1, col=c("red"), cex=.8)
```
```


$$
P[m>200]=1-P[m\leq200]
$$

```{r}
## 2.5%, 50%(i.e. median), 97.5% quantiles of N(161,70)
qnorm(c(0.025, 0.5, 0.975), mean=161, sd=sqrt(70))

## P[m>200]
pnorm(200, mean=161, sd=sqrt(70), lower.tail=FALSE)
```


\begin{itemize}
  \item Plot: see Figure \ref{fig:prior-post}
  \item Summary statistics: see Table \ref{tab:summary}
\end{itemize}

#### 4(c)

$$
\begin{aligned}
y_1,\cdots,y_n\mid m  & \overset{i.i.d}{\sim} \mathcal{N}(m,\kappa^{-1}) \\
m & \: \sim \:\mathcal{N}(\mu,\lambda^{-1}) \\
m\mid y_1,\cdots,y_n & \:\sim\: \mathcal{N}
\left(\frac{\kappa n\bar{y}+\lambda\mu}{n\kappa+\lambda}, (n\kappa+\lambda)^{-1} \right)
\end{aligned}
$$

```{r}
## Posterior mean
((1/900)*2185+(1/70)*161 ) / (13*(1/900)+(1/70) )

## Posterior variance
(13*(1/900)+(1/70))^(-1)
```

$$
\begin{aligned}
y_1,\cdots,y_n &\sim \mathcal{N}(m, 900) \\
m &\sim \mathcal{N}(161,70) \\
m\mid y_1,\cdots,y_n &\sim \mathcal{N}(164.558, 34.80663)
\end{aligned}
$$

```{r, out.width="50%", fig.align='center', fig.cap="Prior vs. Posterior\\label{fig:prior-post}"}
m <- seq(101, 221)
plot(m, dnorm(m, mean=161, sd=sqrt(70)), type="l", ylab="density", col="red", 
     ylim=c(0,0.08), main="Prior vs Posterior")
lines(m, dnorm(m, mean=164.558, sd=sqrt(34.80663)), col="blue")
legend("topright", legend=c("prior: N(161,70)", "posterior: N(164.56, 34.81)"), 
       lty=1, col=c("red", "blue"), cex=.8)
```

```{r}
qnorm(c(0.025, 0.5, 0.975), mean=164.558, sd=sqrt(34.80663))
```

\begin{table}[!htbp]
    \centering
    \caption{Summary statistics of the prior and posterior distributions}
    \begin{tabular}{lcccc}
    \hline
                           & Mean              & SD         & Median   & Equi-tailed $95\%$ CI/CrI \\
    \hline
    $m$                    & $161.000$         & $8.3666$   & $161.000$    & $(144.6018,177.3982)$ \\
    $m\mid y_i,\cdots,y_n$ & $164.558$         & $5.8997$   & $164.558$    & $(152.9948,176.1212)$ \\
    \hline
    \end{tabular}
    \label{tab:summary}
\end{table}

**Interpretation:** there is a posterior probability of 95% that the true mean falls into the interval between 152.9948 and 176.1212, given a $\mathcal{N}(m, 900)$ prior is assumed.

#### 4(d)

$$
P[m>200| y_1, ..., y_n]
$$

```{r}
## P[m>200|y1,...,yn]
pnorm(200, mean=164.558, sd=sqrt(34.80663), lower.tail=FALSE)
```

#### 4(e)

$$
\begin{aligned}
\textrm{Prior} &\to \textrm{Posterior} \\
\mathcal{N}(161,70) &\to \mathcal{N}(164.558, 34.80663) \\
P[m>200] &\to P[m>200\mid y_1,\cdots,y_n] \\
1.570393\times 10^{-6} &\to 9.425552\times 10^{-10}
\end{aligned}
$$

\begin{table}[!htbp]
    \centering
    \caption{Comparision between prior and posterior}
    \begin{tabular}{lcc}
    \hline
                                & Prior                    & Posterior                        \\
    \hline
    Mean                        & $161$                    & $164.558$                        \\
    Variance                    & $70$                     & $34.80663$                       \\
    $\mathcal{N}(\mu,\sigma^2)$ & $\mathcal{N}(161,70)$    & $\mathcal{N}(164.558, 34.80663)$ \\
    $P[m>200]$                  & $1.570393\times 10^{-6}$ & $9.425552\times 10^{-10}$        \\
    \hline
    \end{tabular}
    \label{tab:prior-posterior}
\end{table}

It is noticeable that the variance in particular has decreased from prior to posterior distribution. Thus the dispersion is smaller. This leads to the distribution becoming narrower and the probabilities around the mean increase. Thus, the probability of observing a value greater than 200 also decreases. 

### Exercise 5 (Bayesian learning)

Prior distribution:

$$
p\sim \textrm{Beta}(\alpha,\beta)
$$

Posterior distribution:

$$
p\mid y_1,\cdots,y_n \sim \textrm{Beta}(\alpha+n\bar{y},\beta+n-n\bar{y})
$$

#### 5(a)

$\alpha=\beta=0.5,\quad \textrm{Beta}(0.5,0.5)$


\begin{table}[!htbp]
    \centering
    \caption{Response rate for each stage under $B(0.5,0.5)$ prior}
    \begin{tabular}{lcccc}
    \hline
            & $n$    & Responders       & Prior                      & Posterior                        \\
    Stage   &        & $x\:(\%)$        & $\textrm{B}(\alpha,\beta)$ & $\textrm{B}(\alpha+x,\beta+n-x)$ \\
    \hline
    Interim & $12$   & $3\:(25\%)$      & $\textrm{B}(0.5,0.5)$      & $\textrm{B}(3.5,9.5)$            \\
            &        &                  &                            &                                  \\
    Final   & $64$   & $14\:(21.875\%)$ & $\textrm{B}(0.5,0.5)$      & $\textrm{B}(14.5,50.5)$          \\
    \hline
    \end{tabular}
    \label{tab:tab1}
\end{table}


```{r, figures-side, fig.show="hold", out.width="50%"}
p <- seq(1e-3,1, length=200)

plot(p, dbeta(p, 0.5, 0.5), type="l", ylab="density", col="red", 
     ylim=c(0, 10), main="Intermediate study")
lines(p, dbeta(p, 3.5, 9.5), ylab="density", col="blue")

legend("topright", legend=c("prior: B(0.5,0.5)", "posterior: B(3.5,9.5)"), 
       col=c("red", "blue"), lty=1, cex=.8)

plot(p, dbeta(p, 0.5, 0.5), type="l", ylab="density", col="red", 
     ylim=c(0, 10), main="Final study")
lines(p, dbeta(p, 14.5, 50.5), ylab="density", col="blue")

legend("topright", legend=c("prior: B(0.5,0.5)", "posterior: B(14.5,50.5)"), 
       col=c("red", "blue"), lty=1, cex=.8)
```

$$
P[\textrm{response rate}>0.4]=P[p>0.4]=1-P[p\leq 0.4]
$$

```{r}
# Intermediate study
pbeta(0.4, 3.5, 9.5, lower.tail=FALSE)

# Final study
pbeta(0.4, 14.5, 50.5, lower.tail=FALSE)
```

```{r}
# Intermediate study
## 2.5%, 50%, 97.5% quantiles of Beta(3.5, 9.5)
qbeta(c(0.025, 0.5, 0.975), 3.5, 9.5)

# Final study
## 2.5%, 50%, 97.5% quantiles of Beta(14.5, 50.5)
qbeta(c(0.025, 0.5, 0.975), 14.5, 50.5)
```

\begin{table}[!htbp]
    \centering
    \caption{Summary statistics of posterior distribution ($B(0.5,0.5)$ prior)}
    \begin{tabular}{llccc}
    \hline
                 &                   & Mean       & Median   & $95\%$ CrI          \\
    Stage        & $B(\alpha,\beta)$ &            &          &                     \\
    \hline
    Interim      & $B(3.5,9.5)$      & $0.2692$   & $0.2571$ & $(0.0759, 0.5292)$  \\
                 &                   &            &          &                     \\
    Final        & $B(14.5,50.5)$        & $0.2292$   & $0.2202$ & $(0.1313, 0.3310)$  \\
    \hline
    \end{tabular}
    \label{tab:tab4}
\end{table}

#### 5(b)

$\alpha=8,\beta=24,\quad \textrm{Beta}(8,24)$

\begin{table}[!htbp]
    \centering
    \caption{Response rate for each stage under $B(8,24)$ prior}
    \begin{tabular}{lcccc}
    \hline
            & $n$    & Responders       & Prior                      & Posterior                        \\
    Stage   &        & $x\:(\%)$        & $\textrm{B}(\alpha,\beta)$ & $\textrm{B}(\alpha+x,\beta+n-x)$ \\
    \hline
    Interim & $12$   & $3\:(25\%)$      & $\textrm{B}(8,24)$         & $\textrm{B}(11, 33)$             \\
            &        &                  &                            &                                  \\
    Final   & $64$   & $14\:(21.875\%)$ & $\textrm{B}(8,24)$         & $\textrm{B}(22,74)$              \\
    \hline
    \end{tabular}
    \label{tab:tab1}
\end{table}

```{r, fig.show="hold", out.width="50%"}
p <- seq(1e-3,1, length=200)

plot(p, dbeta(p, 8, 24), type="l", ylab="density", col="red", 
     ylim=c(0, 10), main="Intermediate study")
lines(p, dbeta(p, 11, 33), ylab="density", col="blue")

legend("topright", legend=c("prior: B(8,24)", "posterior: B(11,33)"), 
       col=c("red", "blue"), lty=1, cex=.8)

plot(p, dbeta(p, 8, 24), type="l", ylab="density", col="red", 
     ylim=c(0, 10), main="Final study")
lines(p, dbeta(p, 22, 74), ylab="density", col="blue")

legend("topright", legend=c("prior: B(8,24)", "posterior: B(22,74)"), 
       col=c("red", "blue"), lty=1, cex=.8)
```

$$
P[\textrm{response rate}>0.4]=P[p>0.4]=1-P[p\leq 0.4]
$$

```{r}
# Intermediate study
pbeta(0.4, 11, 33, lower.tail=FALSE)

# Final study
pbeta(0.4, 22, 74, lower.tail=FALSE)
```

```{r}
# Intermediate study
## 2.5%, 50%, 97.5% quantiles of Beta(11, 33)
qbeta(c(0.025, 0.5, 0.975), 11, 33)

# Final study
## 2.5%, 50%, 97.5% quantiles of Beta(22, 74)
qbeta(c(0.025, 0.5, 0.975), 22, 74)
```

\begin{table}[!htbp]
    \centering
    \caption{Summary statistics of posterior distribution ($B(8,24)$ prior)}
    \begin{tabular}{llccc}
    \hline
                 &                   & Mean       & Median   & $95\%$ CrI          \\
    Stage        & Posterior         &            &          &                     \\
    \hline
    Interim      & $B(11,33)$        & $0.2500$   & $0.2461$ & $(0.1352, 0.3863)$  \\
                 &                   &            &          &                     \\
    Final        & $B(22,74)$        & $0.2292$   & $0.2273$ & $(0.1512, 0.3178)$  \\
    \hline
    \end{tabular}
    \label{tab:tab4}
\end{table}
