---
title: "The Beta Distribution"
author: "Group: Goliath"
date: "STA421"
output:
  ioslides_presentation:
    transition: faster
    widescreen: yes
    css: style.css
    logo: uzh_logo.png
  slidy_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, echo=FALSE,include=FALSE}
#load libraries
library(readxl)
library(ggplot2)
library(dplyr,warn.conflicts = FALSE)
library(tidyr)
library(kableExtra)

options(dplyr.summarise.inform = FALSE)
options("kableExtra.html.bsTable" = T)
```


## Properties of the Beta distribution

Continuous density function  defined on the interval $[0,1]$ $f(x)=\frac{1}{B(\alpha,\beta)}x^{\alpha-1}(1-x)^{\beta-1}$ with the two shape parameters $\alpha,\beta>1$

Normalization assured by $B(\alpha,\beta)=\dfrac{\Gamma(\alpha)\Gamma(\beta)}{\Gamma(\alpha+\beta)}=\dfrac{(\alpha-1)!(\beta-1)!}{(\alpha+\beta-1)!}$

Expectation value: $E(X)=\dfrac{\alpha}{\alpha+\beta}$

Variance: $Var(X)=\dfrac{\alpha\beta}{(\alpha+\beta+1)(\alpha+\beta)^2}$

## Properties of the Beta distribution

```{r,echo=F}
x_grid <- seq(from = -0.0, to = 1.0, length.out = 100)
alpha <- c(0.1, 0.5,1, 2,5)
beta <- c(0.1, 0.5,1, 2, 5)

p <- 1.1
p <- 2
alpha <- c(seq(from = 0.0, to = 5, length.out = 20),p)
beta <- c(seq(from = 0.0, to = 5, length.out = 20),p)

par_grid <- expand.grid("alpha"=alpha,"beta"=beta)
y <- data.frame(t(apply(X = par_grid, MARGIN = 1,FUN = function(x){dbeta(x_grid,x[1],x[2])})))
df <- cbind(par_grid,y)
colnames(df)[3:ncol(df)] <- x_grid
df_long <- tidyr::gather(df,x,density,-c(alpha,beta) )
df_long$x <- as.numeric(df_long$x)

ggplot(data = df_long[df_long$alpha==df_long$beta,], aes(x=x, y = density,group = beta,col = beta))+
  geom_line()+
  geom_vline(aes(xintercept = 0.5))

ggplot(data = df_long[df_long$alpha==p,], aes(x=x, y = density,group = beta,col = beta))+
  geom_line()+
  geom_vline(aes(xintercept = 0.5))+
  geom_line(data = df_long[df_long$beta==p & df_long$alpha==p,], aes(x=x, y = density),col = "red")+
  ggtitle(paste0("alpha=",p))

ggplot(data = df_long[df_long$beta==p,], aes(x=x, y = density,group = alpha,col = alpha))+
  geom_line()+
  geom_vline(aes(xintercept = 0.5))+
  ggtitle(paste0("beta=",p))+
  geom_line(data = df_long[df_long$beta==p & df_long$alpha==p,], aes(x=x, y = density),col = "red")
```


```{r,echo=F, include=F}
ggplot(data = df_long[df_long$alpha==sort(unique(df_long$alpha))[2]
,], aes(x=x, y = density,group = beta,col = beta))+
  geom_line()+
  geom_vline(aes(xintercept = 0.5))

ggplot(data = df_long[df_long$beta==sort(unique(df_long$beta))[2]
,], aes(x=x, y = density,group = alpha,col = as.factor(alpha)))+
  geom_line(show.legend = T)
```


## Why is it interesting for Bayesian statistics?
We can change $x$ to $p$ and take it as a prior distribution!



## Reference:

<ul>
  <li>https://en.wikipedia.org/wiki/Beta_distribution#Mean_and_sample_size </li>
</ul>
