---
title: "Worksheet 2"
subtitle: "Foundations of Bayesian Methodology"
author: "Goliath"
date: "Spring Semester 2022"
output: 
  html_document:
    toc: true
    toc_depth: 5
    toc_float: true
# runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("F:/UZH/22Spring/FBM/Session3/R")
Sys.setenv(lang="us_en")
rm(list=ls())
```

### Group Contribution

**Task:** Write a short program in $\texttt{R}$ to estimate the area of the unit circle $x^2+y^2=1$ by randomly dropping points (independent Monte Carlo sampling) on it.

The program proceeds as follows:

* Simulate a large number of points uniformly distributed in the square $[-1,1]\times[-1,1]$.
* Calculate the distance from the origin to each point.
* Calculate the ratio of the number of points with a distance less than 1 to the total number of points simulated.
* This ratio is nothing else than the estimate of the area of the unit circle divided by the area of the square.
* The estimated area of the unit circle is simply the ratio times the area of the square.

$$
\frac{\textrm{area of unit circle}}{\textrm{area of square}}
\approx\frac{\textrm{# of points within the circle}}{\textrm{# of points simulated}}
$$

$~$

#### Implementation with a function in R

```{r libraries, message=FALSE, warning=FALSE}
## Load libraries
library(dplyr)
library(ggplot2)
library(gganimate)
library(gifski)
library(shiny)
```

```{r}
## Set seed for reproducible results
set.seed(2022)

## Define a function
## Argument:
##   n: the number of points in MC simulation

estCircleArea <- function(n = 1000) {
  ## Library dependencies
  library(dplyr)
  library(ggplot2)
  
  ## Generate random coordinates within the square
  d.points <- data.frame(
    x = runif(n, min=-1, max=1),
    y = runif(n, min=-1, max=1)
  )
  
  ## Adding two columns to the data frame
  d.points <- d.points %>% mutate(
    distance = sqrt(x^2 + y^2),                       # add distance variable
    withinCircle = ifelse(distance <= 1, TRUE, FALSE) # add boolean variable
  )
  
  ## Estimate the area
  ratio <- mean(d.points$withinCircle == TRUE) 
  # points within circle : total points ~ circle area : square area
  
  area <- ratio * 4 # the estimated area of unit circle
  
  ## Generate plot
  plot <- ggplot(d.points) + 
    geom_point(aes(x=x, y=y, color=withinCircle)) +
    coord_fixed(ratio=1) + 
    theme_minimal()
  
  return(list(plot=plot, area=area))
}
```

$~$

#### Visualization with animated plots

```{r, fig.align="center"}
## Unit Circle

n    <- 5000     # the number of points generated
step <- 100      # the points generated with a step size of 100 per frame

## Generate random coordinates within the square
d.points <- data.frame(
  points = 1:n, 
  x = runif(n, min=-1, max=1),
  y = runif(n, min=-1, max=1)
)

## Adding two columns to the data frame
d.points <- d.points %>% mutate(
  run = rep(1:(n/step), each=step), 
  distance = sqrt(x^2 + y^2),                       # add distance variable
  withinCircle = ifelse(distance <= 1, TRUE, FALSE) # add boolean variable
)

p <- ggplot(d.points) + 
  geom_point(aes(x=x, y=y, color=withinCircle)) +
  coord_fixed(ratio=1) + 
  theme_minimal() + 
  transition_states(run) + 
  shadow_mark() + 
  ggtitle("Monte Carlo Simulation")

animate(p, 
        width = 600, # 600px wide
        height = 600, # 600px high
        duration = 20,
        renderer = gifski_renderer())
```


```{r, fig.align="center"}
## Traceplot
dat <- as.data.frame(matrix(nrow=500, ncol=2))
colnames(dat) <- c("Iteration", "Estimate")

for (i in 1:500) {
  est <- estCircleArea(n = i*100)
  dat[i, 1] <- i
  dat[i, 2] <- est$area
}

traceplot <- ggplot(dat, aes(x=Iteration, y=Estimate)) + 
  geom_line(color=3) + 
  geom_point(color=3) + 
  geom_hline(yintercept=pi, color=2, linetype="longdash") +
  geom_text(aes(0, pi, label=pi, vjust=1.2, hjust=0.2), color=2) + 
  labs(title="Traceplot", x="Number of points / 100", y="Estimation") + 
  theme_minimal() + 
  transition_reveal(Iteration)

animate(traceplot, 
        width = 600, # 600px wide
        height = 400, # 400px high
        duration = 20,
        renderer = gifski_renderer())
```

$~$

#### Visualization with interactive Shiny

```{r interactive web app, eval=FALSE}
# User interface with navigation bar
ui <- pageWithSidebar(
  ## Header
  headerPanel("Monte Carlo Sampling"),
  
  ## Getting the number of trials
  sidebarPanel(
    sliderInput("N",
                "Number of random points:",
                min = 1000,
                max = 19000,
                step = 1000,
                value = 1000),
  ),
  
  ## Displaying the output
  mainPanel(plotOutput("plot1"), textOutput("text1"))
)


# Server logic
server <- function(input, output) {
  
  ## Monte Carlo simulation
  output$plot1 <- renderPlot({
    ## Set seed for reproducible results
    set.seed(2022)
    trials <- input$N
    
    ## Generate random coordinates within the square
    d.points <- data.frame(
      x = runif(trials, min=-1, max=1),
      y = runif(trials, min=-1, max=1)
    )
    
    ## Adding two columns to the data frame
    d.points <- d.points %>% mutate(
      distance = sqrt(x^2 + y^2),                       # add distance variable
      withinCircle = ifelse(distance <= 1, TRUE, FALSE) # add boolean variable
    )
    
    ## Generate plot
    ggplot(d.points) + 
      geom_point(aes(x=x, y=y, color=withinCircle)) +
      coord_fixed(ratio=1) + 
      theme_minimal()
  })
  
  output$text1 <- renderText({
    
    ## Set seed for reproducible results
    set.seed(2022)
    trials <- input$N
    
    ## Generate random coordinates within the square
    d.points <- data.frame(
      x = runif(trials, min=-1, max=1),
      y = runif(trials, min=-1, max=1)
    )
    
    ## Adding two columns to the data frame
    d.points <- d.points %>% mutate(
      distance = sqrt(x^2 + y^2),                       # add distance variable
      withinCircle = ifelse(distance <= 1, TRUE, FALSE) # add boolean variable
    )
    
    ## Estimate the area
    ratio <- mean(d.points$withinCircle == TRUE) 
    
    ## the estimated area of unit circle
    area <- ratio * 4 
    paste("The area of the unit circle", area)
  })
}

shinyApp(ui = ui, server = server)
```

