---
title: "Worksheet 2"
author: "Goliath"
date: "Spring Semester 2022"
output:
  html_document:
    toc: yes
    toc_depth: 5
    toc_float: yes
subtile: Foundations of Bayesian Methodology
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
Sys.setenv(lang="us_en")
rm(list=ls())
```

### Group Contribution

Write a short program in \texttt{R} to estimate the area of the unit circle $x^2+y^2=1$ by dropping randomly points (independent Monte Carlo sampling) on it.

* Simulate a large number of points uniformly distributed in the square $[-1,1]\times[-1,1]$.
* Check the distance of the point from the origin.
* Calculate the ratio of the number of points with a distance from origin less than 1 to the total number of points simulated.
* The ratio is an estimate of the ratio of the area of the unit circle to the area of the square.

$$
\frac{\textrm{area of unit circle}}{\textrm{area of square}}
\approx\frac{\textrm{# of points within the circle}}{\textrm{# of total points}}
$$

#### Creating a Function

```{r, warning=FALSE, message=FALSE}
# Define a function
## Parameter:
##   n: the number of points in MC simulation

estCircleArea <- function(n = 1000) {
  ## Library dependencies
  library(dplyr)
  library(ggplot2)
  
  ## Generate random coordinates within the square
  d.points <- data.frame(
    x = runif(n, min=-1, max=1),
    y = runif(n, min=-1, max=1)
  )
  
  ## Adding two columns to the data frame
  d.points <- d.points %>% mutate(
    radius = sqrt(x^2 + y^2),                       # add radius variable
    withinCircle = ifelse(radius <= 1, TRUE, FALSE) # add boolean variable
  )
  
  ## Generate plot
  plot <- ggplot(d.points) + 
    geom_point(aes(x=x, y=y, color=withinCircle)) +
    coord_fixed(ratio=1) + 
    theme_minimal()
  
  ## Estimate the area
  ratio <- mean(d.points$withinCircle == TRUE) 
  # points within circle : total points ~ circle area : square area
  
  area <- ratio * 4 # the estimated area of unit circle
  
  # cat(sprintf("The estimated area of the unit circle is %.4f", area))
  return(list(plot=plot, area=area))
}
```

#### Visualization

```{r, warning=FALSE, message=FALSE}
library(ggplot2)
library(gganimate)
library(gifski)
```

```{r, message=FALSE, warning=FALSE}
dat <- as.data.frame(matrix(nrow=500, ncol=2))
colnames(dat) <- c("Iteration", "Estimate")

for (i in 1:500) {
  est <- estCircleArea(n = i*100)
  dat[i, 1] <- i
  dat[i, 2] <- est$area
}

ggplot(dat, aes(x=Iteration, y=Estimate)) + 
  geom_line(color=3) + 
  geom_point(color=3) + 
  geom_hline(yintercept=pi, color=2, linetype="longdash") +
  theme_minimal()

p <- ggplot(dat, aes(x=Iteration, y=Estimate), 
            label=paste("Estimate:", format())) + 
  geom_line(color=3) + 
  geom_point(color=3) + 
  geom_hline(yintercept=pi, color=2, linetype="longdash") +
  theme_minimal() + 
  transition_reveal(Iteration)

animate(p, # fps=50, 
        width = 600, # 900px wide
        height = 400, # 600px high
        duration = 20,
        renderer = gifski_renderer())
```


```{r}
n <- 5000
## Generate random coordinates within the square
d.points <- data.frame(
  points = 1:n, 
  x = runif(n, min=-1, max=1),
  y = runif(n, min=-1, max=1)
)

## Adding two columns to the data frame
d.points <- d.points %>% mutate(
  run = rep(1:50, each=100), 
  radius = sqrt(x^2 + y^2),                       # add radius variable
  withinCircle = ifelse(radius <= 1, TRUE, FALSE) # add boolean variable
)
```

```{r}
ggplot(d.points, aes(x=x, y=y, color=withinCircle)) + 
  geom_point() +
  coord_fixed(ratio=1) + 
  theme_minimal()

p <- ggplot(d.points) + 
  geom_point(aes(x=x, y=y, color=withinCircle)) +
  coord_fixed(ratio=1) + 
  theme_minimal() + 
  transition_states(run) + 
  shadow_mark()

animate(p, # fps=50, 
        width = 400, # 900px wide
        height = 400, # 600px high
        duration = 20,
        renderer = gifski_renderer())
```


```{r, warning=FALSE, message=FALSE}
## Set seed for reproducible results
set.seed(2022)

## Visualization
est <- estCircleArea(n = 5000)

est$area
est$plot
```

#### Traceplot

```{r}
library(gganimate)
library(gapminder)
library(gifski)
dat <- as.data.frame(matrix(nrow=500, ncol=2))
colnames(dat) <- c("Iteration", "Estimate")

for (i in 1:500) {
  est <- estCircleArea(trials = i*100)
  dat[i, 1] <- i
  dat[i, 2] <- est$area
}

p <- ggplot(dat, aes(x=Iteration, y=Estimate)) + 
  geom_line() + 
  geom_point() + 
  transition_reveal(Iteration)

anim_save("demo.gif")
```

```{r}
ggplot(dat, aes(x=Iteration, y=Estimate)) + 
  geom_line(color=3) + 
  geom_point(color=3) + 
  geom_hline(yintercept=pi, color=2, linetype="longdash",) + 
  labs(x="Iteration with a step size of 100", y="Estimate", title="Monte Carlo") +
  theme_minimal()
```


```{r}
library(gapminder)
library(ggplot2)
library(gganimate)
data(gapminder)
p <- ggplot(
gapminder,
aes(x = gdpPercap, y=lifeExp, size = pop, colour = country)
) +
geom_point(show.legend = FALSE, alpha = 0.7) +
scale_color_viridis_d() +
scale_size(range = c(2, 12)) +
scale_x_log10() +
labs(x = "GDP per capita", y = "Life expectancy")
p + transition_time(year)
```


```{r}
## Set seed for reproducible results
set.seed(2022)

## Number of simulations
iterations <- 500 

## Initialize a vector to store estimated area for unit circle
vec.area <- as.numeric(iterations)

for (i in 1:iterations) {
  est <- estCircleArea(trials = i*100)
  vec.area[i] <- est$area
}

## Traceplot
plot(1:iterations, vec.area, type="l", ylab="Area of unit circle",
     xlab="Iterations with a step size of 100", main="Traceplot")
abline(h=pi, lty=3, col="red", lwd=2)
```

```{r}
## Library dependencies
library(dplyr)
library(ggplot2)

dat <- as.data.frame(matrix(nrow=55000, ncol=6))
colnames(dat) <- c("points", "x", "y", "radius", "withinCricle")

for (i in 1:10) {
  for (j in )
}

for (i in 1:4) {
  n.points <- i * 1000 # number of points

  ## Generate random coordinates within the square
  d.points <- data.frame(
    x = runif(n.points, min=-1, max=1),
    y = runif(n.points, min=-1, max=1)
  )
  
  ## Adding two columns to the data frame
  d.points <- d.points %>% mutate(
    radius = sqrt(x^2 + y^2),                       # add radius variable
    withinCircle = ifelse(radius <= 1, TRUE, FALSE) # add Boolean variable
  )
  
  ## Generate plot
  plot <- ggplot(d.points) + 
    geom_point(aes(x=x, y=y, color=withinCircle)) +
    coord_fixed(ratio=1) + 
    theme_minimal()
  return(plot)
}


```

