---
title: "Worksheet01_Exercise06"
author: "Goliath"
date: "25 2 2022"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Exercise 6 - Monte Carlo: Random Sample vs. the True Distribution

```{r, include=T}
set.seed(44566)
X <- rnorm(n=1000, mean = 160, sd = 20)
```



## Part 1 - True Values:

The expected value of the standard normal distribution equals 0: 
$$X \overset{}{\sim} Norm(\mu = 0, \sigma^2= 1)$$ It follows that $E(X) = 0$.

When $Y \overset{}{\sim} Norm(\mu, \sigma^2)$ then $$X = \frac{(Y-\mu)}{\sigma}$$ is standard normally distributed. Therefore $$E(Y) = E(\sigma X+\mu) = \sigma E(X)+ \mu = \mu$$

The true values (= theoretical) of the distribution are the following:


```{r, include=T}
### Calculations:
mean_t <- 160
sd_t <- 20
var_t <- sd_t^2
med_t <- 160
q0025_t <- qnorm(0.025, 160,20) #120.8007
q50_t <- qnorm(0.5, 160, 20)  #160
q975_t <- qnorm(0.975,160,20)  #199.1993
True <- c(mean_t, sd_t, var_t, med_t, round(q0025_t,2), 
                round(q50_t,2), round(q975_t,2))

values<- as.data.frame(row.names = c("mean", "sd", "variance", "median", "2.5% Quantile", 
                "50% Quantile", "97.5% Quantile"),
                x = True)

## True Values:
values
```


## Part 2 - Traceplot:

```{r, echo=T}
df <- as.data.frame(X)
df$m <- seq(1,1000,1)

plot(df$m, df$X, type = "l", main="Traceplot", xlab = "Iteration", ylab = "X")
```


## Part 3 - Histogram and True Density Curve:

```{r}
hist(X, xlim = c(75,250), freq = F, ylim = c(0,0.02), main = "Histogram of Sample and True Density Curve")
lines(dnorm(0:250,mean=160,sd=20), col="red", lwd=2)

```


## Part 4 - Sample Values:


```{r, include=T}
### Calculations:
mean_hat <- mean(X)
sd_hat <- sd(X)
var_hat <- var(X)
med_hat <- median(X)
q0025_hat <- quantile(X, probs = 0.025)[[1]]
q50_hat <- quantile(X, probs = 0.5)[[1]]
q975_hat <- quantile(X, probs = 0.975)[[1]]

sample_values <- c(round(mean_hat,2), round(sd_hat,2), round(var_hat,2), 
                   round(med_hat,2), round(q0025_hat,2), 
                   round(q50_hat,2), round(q975_hat,2))

values$Sample <- sample_values

## 
values
```

```{r, include=T}
## Comparision:
values$Difference <- values$True- values$Sample
values
```


## Part 5 - Probabilities:


```{r, include=T}
## P[X>175] = 1-P[X <= 175]:
## Probability of the Sample: 
p1 <- length(which(X>175))/1000
p1

## Theoretical Probability:
1- pnorm(175, mean = 160, sd = 20)

## P[150 < X < 180] = P[151 <= X <= 179]
## Probability of the Sample: 
p2 <- length(which(X>150&X<180))/1000
p2

## Theoretical Probability:
pnorm(179, 160, 20) - pnorm(151, 160, 20)

```
Comment:

$P[X > 175]$ is equal to $1-P[X \le 175]$. 

$P[150 < X < 180]$ is equal to $P[X \le 179] - P[X \le 151]$. 


